Inference of Gravitational Lens Time Delays Using Gaussian Processes

### 2013-05-03 doc/Leon/ms.tex
Much more writing

\section{Model Assumptions}\label{sec:assumptions}
Suppose that the underlying QSO variability as a function
of time, as seen through image $A$, is described by the following function:
\begin{eqnarray}
y_{\rm QSO}(t)
\end{eqnarray}
Then the light curves of the two images
$A$ and $B$ will be given by:
\begin{eqnarray}
y_A(t) &=& M_A + y_{\rm QSO}(t)\\
y_B(t) &=& M_B + y_{\rm QSO}(t - \tau).
\end{eqnarray}
The light curve of image $A$ is just the underlying QSO variability,
and the light curve of

plus a signal $\mu_A(t)$ that is due to microlensing or perhaps
other effects. The variability of image $B$ is given by
a lagged version of the QSO variability, plus another
perturbation $\mu_B(t)$ caused my microlensing.

The data are assumed to be noisy measurements of $y_A(t)$ and
$y_B(t)$ at various times.

If we make the simplifying assumption that our prior knowledge about
$\mu_A(t)$, $\mu_B(t)$, and
$y_{\rm QSO}(t)$ can be described by Gaussian Processes \citep{rasmussen},
then the probability distribution for the data $\yy$ given the parameters
$\theta$
is a multivariate normal distribution, where the mean vector $\mm$
and the covariance matrix $\CC$ depend on the parameters:
\begin{eqnarray}
p(\yy | \theta) &=& \frac{1}{\sqrt{(2\pi)^N\det{\CC}}}
\exp\left[
-\frac{1}{2}
\left(
\yy - \mm
\right)^T
\CC^{-1}
\left(
\yy - \mm
\right)
\right]
\end{eqnarray}
The elements of the mean vector are just the mean magnitudes of the
appropriate images:
\begin{eqnarray}
m_i = M_{c_i}
\end{eqnarray}

The covariance matrix elements are given by:
\begin{eqnarray}
C_{ij} &=& \textnormal{Cov}(y_i, y_j) \\
&=& \mathds{E}\left[
\left(y_i - m_i\right)
\left(y_j - m_j\right)
\right]
\end{eqnarray}

We assume the following:
\begin{eqnarray}
\textnormal{Cov}\left(y_{\rm QSO}(t_1), y_{\rm QSO}(t_2)\right)
&=& \sigma_{\rm QSO}^2\exp\left[-\frac{\left|t_2 - t_1\right|}{L_{\rm QSO}}\right]
\end{eqnarray}

\begin{eqnarray}
\textnormal{Cov}\left(\mu_A(t_1), \mu_A(t_2)\right)
&=& \sigma_{{\rm ML}, A}^2\exp\left[-\left(\frac{\left|t_2 - t_1\right|}{L_{{\rm ML}, A}}\right)^\alpha\right]
\end{eqnarray}

\begin{eqnarray}
\textnormal{Cov}\left(\mu_B(t_1), \mu_B(t_2)\right)
&=& \sigma_{{\rm ML}, B}^2\exp\left[-\left(\frac{\left|t_2 - t_1\right|}{L_{{\rm ML}, B}}\right)^\alpha\right]
\end{eqnarray}

Numerically, the logarithm of the determinant and the
$\CC^{-1}\left(\yy - \mm\right)$ term are implemented using
Cholesky decompositions.

\begin{table}[h!]
\begin{center}
\begin{tabular}{|c|c|c|}
\hline
Parameter & Description & Prior\\
\hline
$\{m_k\}$ & Mean magnitudes & Uniform$(m_{\rm min}, m_{\rm max})$\\
$\{\tau_k\}$ & Time delays & Cauchy$(0, \tau_{\rm range}/10)T(\tau_{\rm min}, \tau_{\rm max})$\\
$\{\sigma_{\rm ML}\}$ & Standard deviations of microlensing signals & LogUniform$(\sigma_{\rm min}, \sigma_{\rm max})$\\
$\{L_{\rm ML}\}$ & Timescales of microlensing signals & LogUniform$(L_{\rm min}, L_{\rm max})$\\
$\alpha$ & Smoothness parameter for microlensing signals & Uniform$(1,2)$\\
$\sigma_{\rm QSO}$ & Standard deviation of quasar variability & LogUniform$(\sigma_{\rm min}, \sigma_{\rm max})$\\
$L_{\rm QSO}$ & Timescale of quasar variability & LogUniform$(L_{\rm min}, L_{\rm max})$\\
$f_{\rm bad}$ & Fraction of data points with incorrect errorbars & Uniform$(0,1)$\\
$K$ & Boost factor for bad errorbars & LogUniform$(1, 100)$\\
$\{b_i\}$ & Flag for whether each errorbar is bad & Bernoulli$(f_{\rm bad})$\\
\hline
\end{tabular}
\end{center}
\end{table}

\begin{figure}
\begin{center}
\includegraphics[scale=0.5]{hist.pdf}
\end{center}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[scale=0.5]{error_bars.pdf}
\end{center}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[scale=0.7]{posterior_predictive_check.pdf}
\end{center}
\end{figure}

\begin{thebibliography}{99} 
\bibitem[\protect\citeauthoryear{Rasmussen \& Williams}{2006}]{rasmussen} Rasmussen C., Williams C., Gaussian Processes for Machine Learning, The MIT Press, 2006
\end{thebibliography}

>>>>

\section{Model Assumptions}\label{sec:assumptions}
Suppose that the underlying QSO variability as a function
of time, as seen through image $A$, is described by the following function:
\begin{eqnarray}
y_{\rm QSO}(t)
\end{eqnarray}
Then the light curves of the two images
$A$ and $B$ will be given by:
\begin{eqnarray}
y_A(t) &=& M_A + y_{\rm QSO}(t) + \mu_A(t)\\
y_B(t) &=& M_B + y_{\rm QSO}(t - \tau) + \mu_B(t).
\end{eqnarray}
The light curve of image $A$ is just the underlying QSO variability,
fluctuating around the mean magnitude $M_A$ of image $A$, plus
a microlensing signal $\mu_A(t)$. Similarly,
the light curve of image $B$ is its mean magnitude $M_B$ plus the QSO
variability delayed by a lag $\tau$, plus a different microlensing
signal $\mu_B(t)$. The data are assumed to be noisy measurements of $y_A(t)$ and
$y_B(t)$ at various times.

\subsection{Error Bar Wrongness}
We do not assume that the given error bars
$\{\sigma_i\}_{i=1}^N$ are entirely correct. Instead, we assume that each
measurement is either ``good'', in which case $\sigma_i$ is used, otherwise it
is ``bad'', in which case $K\sigma_i$ is used, where $K > 1$. Letting $b_i = 1$
denote a bad measurement and $b_i=0$ denote a good measurement, the PDF for the
data can be written as:
\begin{eqnarray}
y_i \sim \mathcal{N}
\left(
y_{c_i}(t_i), \sigma_i'^2
\right)\label{eq:noise}
\end{eqnarray}
where
\begin{eqnarray}
\sigma_i' &=& \left\{
\begin{array}{lrl}
\sigma_i, & b_i = 1\\
K\sigma_i, & b_i = 0 & .
\end{array}
\right.
\end{eqnarray}

\section{Unknown Parameters}
In Table~\ref{tab:params} we list the unknown parameters that are to be inferred
and the prior distributions that were used.

\begin{table}[h!]
\begin{center}
\begin{tabular}{|c|c|c|}
\hline
Parameter & Description & Prior\\
\hline
$\{M_A, M_B\}$ & Mean magnitudes of the two images & Uniform$(M_{\rm min}, M_{\rm max})$\\
$\tau$ & Time delay & Cauchy$(0, \tau_{\rm range}/10)T(\tau_{\rm min}, \tau_{\rm max})$\\
$\{\sigma_{{\rm ML}, A}, \sigma_{{\rm ML}, B}\}$ & Standard deviations of microlensing signals & LogUniform$(\sigma_{\rm min}, \sigma_{\rm max})$\\
$\{L_{{\rm ML, A}}, L_{{\rm ML, B}}\}$ & Timescales of microlensing signals & LogUniform$(L_{\rm min}, L_{\rm max})$\\
$\alpha$ & Smoothness parameter for microlensing signals & Uniform$(1,2)$\\
$\sigma_{\rm QSO}$ & Standard deviation of quasar variability & LogUniform$(\sigma_{\rm min}, \sigma_{\rm max})$\\
$L_{\rm QSO}$ & Timescale of quasar variability & LogUniform$(L_{\rm min}, L_{\rm max})$\\
$f_{\rm bad}$ & Fraction of data points with incorrect errorbars & Uniform$(0,1)$\\
$K$ & Boost factor for bad errorbars & LogUniform$(1, 100)$\\
$\{b_i\}_{i=1}^N$ & Flag for whether each errorbar is bad & Bernoulli$(f_{\rm bad})$\\
\hline
\end{tabular}
\end{center}
\caption{A list of the model parameters and their prior probability
distributions. All of the priors are vague.\label{tab:params}}
\end{table}

\section{Gaussian Processes}
Inspection of Equation~\ref{eq:noise} suggests that this is a simple curve
fitting problem. However, this is not the case, because the noise-free
model predictions $y_{c_i}(t_i)$ cannot be computed as a function of the
parameters listed in Table~\ref{tab:params}.

If we make the simplifying assumption that our prior knowledge about
$\mu_A(t)$, $\mu_B(t)$, and
$y_{\rm QSO}(t)$ can be described by Gaussian Processes \citep{rasmussen},
then the probability distribution for the data $\yy$ given the parameters
$\theta$
is a multivariate normal distribution, where the mean vector $\mm$
and the covariance matrix $\CC$ depend on the parameters:
\begin{eqnarray}
p(\yy | \theta) &=& \frac{1}{\sqrt{(2\pi)^N\det{\CC}}}
\exp\left[
-\frac{1}{2}
\left(
\yy - \mm
\right)^T
\CC^{-1}
\left(
\yy - \mm
\right)
\right]\label{eq:likelihood}
\end{eqnarray}
The Gaussian Process assumption is questionable in the case of the microlensing
signals $\mu_A(t)$ and $\mu_B(t)$ (especially if the source is not large),
although it is still a flexible model for any temporally-correlated signal.
The use of a Gaussian Process model for the QSO variability is more
well established \citep{kelly, xu}.

The elements of the mean vector are just the mean magnitudes of the
appropriate images:
\begin{eqnarray}
m_i = M_{c_i}
\end{eqnarray}
The covariance matrix elements are given by:
\begin{eqnarray}
C_{ij} &=& \textnormal{Cov}(y_i, y_j) \\
&=& \mathds{E}\left[
\left(y_i - m_i\right)
\left(y_j - m_j\right)
\right]
\end{eqnarray}
and are obtained by evaluating the following covariance {\it functions}
at the relevant time differences.
\begin{eqnarray}
\textnormal{Cov}\left(y_{\rm QSO}(t_i), y_{\rm QSO}(t_j)\right)
&=& \sigma_{\rm QSO}^2\exp\left[-\frac{\left|t_i - t_j\right|}{L_{\rm QSO}}\right]\\
\textnormal{Cov}\left(\mu_A(t_i), \mu_A(t_j)\right)
&=& \sigma_{{\rm ML}, A}^2\exp\left[-\left(\frac{\left|t_i - t_j\right|}{L_{{\rm ML}, A}}\right)^\alpha\right]\\
\textnormal{Cov}\left(\mu_B(t_i), \mu_B(t_j)\right)
&=& \sigma_{{\rm ML}, B}^2\exp\left[-\left(\frac{\left|t_i - t_j\right|}{L_{{\rm ML}, B}}\right)^\alpha\right]
\end{eqnarray}
Numerically, the log of the likelihood function (Equation~\ref{eq:likelihood})
requires the log of $\det(\mathbf{C})$ and the
$\CC^{-1}\left(\yy - \mm\right)$ term. These can be implemented using
Cholesky decompositions. Unfortunately the evaluation of this likelihood
function takes $O(N^3)$ time, but is tractable for the B1600 data where $N=604$.

\section{Results}
The MCMC was run using the Diffusive Nested Sampling algorithm
\citep[][software available at {\tt http://github.com/eggplantbren/DNest3}]{dnest}.
The posterior PDF for the time delay is shown in Figure~\ref{fig:time_delay}.

\begin{figure}
\begin{center}
\includegraphics[scale=0.5]{hist.pdf}
\caption{The posterior PDF for the time delay.\label{fig:time_delay}}
\end{center}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[scale=0.5]{error_bars.pdf}
\caption{The joint posterior PDF for the fraction of ``bad'' error bars
and the error bar scaling parameter $K$. The dotted lines correspond to the
assumption that the reported error bars are correct.\label{fig:time_delay}}
\end{center}
\end{figure}

\section{Posterior Predictive Check}
A useful sanity check for a Bayesian model is to verify that the observed data
is typical of the kind of data that the model predicts. This is formalised
using ``posterior predictive checks'' \citep{gelman}. Figure~\ref{fig:ppc}
shows the actual data along with eight simulated datasets.

\begin{figure}
\begin{center}
\includegraphics[scale=0.7]{posterior_predictive_check.pdf}
\caption{Eight data sets simulated from the model
using parameters drawn from the posterior distribution. The actual data set
is also plotted and appears in the lower left panel. The actual data appears
typical of the distribution of simulated data. The most atypical feature of
the real data is the rapid variability in the final season, although similar
features can be seen in the first two simulated datasets.
\label{fig:ppc}}
\end{center}
\end{figure}

\begin{thebibliography}{99}

\bibitem[\protect\citeauthoryear{Brewer, P{\'a}rtay,
\& Cs{\'a}nyi}{2011}]{dnest} Brewer B.~J., P{\'a}rtay L.~B.,
Cs{\'a}nyi G., 2011, Statistics and Computing, 21, 4, 649-656. arXiv:0912.2380

\bibitem[\protect\citeauthoryear{Gelman, Meng, \& Stern}{1996}]{gelman} Gelman, A.,
Meng X., Stern, H., Posterior predictive assessment of model fitness via realized discrepancies,
Statistica Sinica 6 (1996): 733-759.

\bibitem[Kelly et al.(2009)]{kelly} Kelly, B.~C., Bechtold, 
J., Siemiginowska, A.\ 2009.\ Are the Variations in Quasar Optical Flux 
Driven by Thermal Fluctuations?.\ The Astrophysical Journal 698, 895-910.

\bibitem[\protect\citeauthoryear{Rasmussen \& Williams}{2006}]{rasmussen} Rasmussen C., Williams C., Gaussian Processes for Machine Learning, The MIT Press, 2006

\bibitem[Zu et al.(2013)]{zu} Zu, Y., Kochanek, C.~S., 
Koz{\l}owski, S., Udalski, A.\ 2013.\ Is Quasar Optical Variability a 
Damped Random Walk?.\ The Astrophysical Journal 765, 106. 
\end{thebibliography}
